---
title: "Advanced Reactivity"
subtitle: "rstudio::conf 2017"
author: "Joe Cheng"
date: "2017/01/11"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{css}
.callout-bold strong {
  font-weight: normal;
  color: #444444;
  background-color: #FFFFAA;
}
.hljs-keyword {
  font-weight: normal !important;
}
th {
  text-align: left;
}
td, th {
  padding: 3px 20px 3px 0;
}
```

```{r echo=FALSE}
library(shiny)
```

## Reactivity catalog

Store values: [`reactiveValues`](http://shiny.rstudio.com/reference/shiny/latest/reactiveValues.html) / `input` / [`makeReactiveBinding`](http://shiny.rstudio.com/reference/shiny/latest/makeReactiveBinding.html)

Calculate values: [`reactive`](http://shiny.rstudio.com/reference/shiny/latest/reactive.html) / [`eventReactive`](http://shiny.rstudio.com/reference/shiny/latest/observeEvent.html)

Execute tasks: [`observe`](http://shiny.rstudio.com/reference/shiny/latest/observe.html) / [`observeEvent`](http://shiny.rstudio.com/reference/shiny/latest/observeEvent.html)

Preventing reactivity: [`isolate`](http://shiny.rstudio.com/reference/shiny/latest/isolate.html)

Checking preconditions: [`req`](http://shiny.rstudio.com/reference/shiny/latest/req.html)

Time (as a reactive source): [`invalidateLater`](http://shiny.rstudio.com/reference/shiny/latest/invalidateLater.html) / [~~`reactiveTimer`~~](http://shiny.rstudio-staging.com/reference/shiny/latest/reactiveTimer.html)

Rate-limiting: [`debounce`](http://shiny.rstudio-staging.com/reference/shiny/latest/debounce.html) / [`throttle`](http://shiny.rstudio-staging.com/reference/shiny/latest/debounce.html)

Live data: [`reactiveFileReader`](http://shiny.rstudio-staging.com/reference/shiny/latest/reactiveFileReader.html) / [`reactivePoll`](http://shiny.rstudio-staging.com/reference/shiny/latest/reactivePoll.html)

(Pretty sure this is just the beginning...)

---
class: callout-bold

## Reactivity catalog

Store values: [**`reactiveValues`**](http://shiny.rstudio.com/reference/shiny/latest/reactiveValues.html) / `input` / [`makeReactiveBinding`](http://shiny.rstudio.com/reference/shiny/latest/makeReactiveBinding.html)

Calculate values: [**`reactive`**](http://shiny.rstudio.com/reference/shiny/latest/reactive.html) / [`eventReactive`](http://shiny.rstudio.com/reference/shiny/latest/observeEvent.html)

Execute tasks: [**`observe`**](http://shiny.rstudio.com/reference/shiny/latest/observe.html) / [`observeEvent`](http://shiny.rstudio.com/reference/shiny/latest/observeEvent.html)

Preventing reactivity: [**`isolate`**](http://shiny.rstudio.com/reference/shiny/latest/isolate.html)

Checking preconditions: [**`req`**](http://shiny.rstudio.com/reference/shiny/latest/req.html)

Time (as a reactive source): [**`invalidateLater`**](http://shiny.rstudio.com/reference/shiny/latest/invalidateLater.html) / [~~`reactiveTimer`~~](http://shiny.rstudio-staging.com/reference/shiny/latest/reactiveTimer.html)

Rate-limiting: [`debounce`](http://shiny.rstudio-staging.com/reference/shiny/latest/debounce.html) / [`throttle`](http://shiny.rstudio-staging.com/reference/shiny/latest/debounce.html)

Live data: [`reactiveFileReader`](http://shiny.rstudio-staging.com/reference/shiny/latest/reactiveFileReader.html) / [`reactivePoll`](http://shiny.rstudio-staging.com/reference/shiny/latest/reactivePoll.html)

(Pretty sure this is just the beginning...)

**Highlighted functions are fundamental**, all others are built on top.

---

## Review: Reactive expressions

Use to *calculate new values* based on reactive values and other reactive expressions.

*Caches* its return value, until notified of reactive dependencies being out-of-date.

*Lazily* executes--Shiny wants to avoid running these whenever possible. For this reason, meaningful *side effects* are prohibited from reactive expressions.

Call it like a function when you want to read its value.

---

## Review: Reactive expressions

```{r eval=FALSE}
# Declare
movies_subset <- reactive({
  movies %>% filter(title_type %in% input$type)
})

# Read
output$scatterplot <- renderPlot({
  ggplot(movies_subset(), aes(...)) + geom_point()
})
```

---

## Review: Observers

Use to *execute actions* based on changing reactive values and other reactive expressions.

Doesn't return a value. So performing *side effects* is usually the only reason you'd want to create one of these.

*Eagerly* executed by Shiny.

```{r eval=FALSE}
observe({
  print(paste("The value of x is", input$x))
})

## [1] The value of x is 10
## [1] The value of x is 16
## [1] The value of x is 9
```

---

## Reactive expressions vs observers

<table>
<tr><th>reactive()</th><th>observe()</th></tr>
<tr><td>Callable</td><td>Not callable</td></tr>
<tr><td>Returns a value</td><td>No return value</td></tr>
<tr><td>Lazy</td><td>Eager</td></tr>
<tr><td>Cached</td><td>N/A</td></tr>
<tr><td>No side effects</td><td>Only for side effects</td></tr>
</table>

---

## Reactive expressions vs observers vs functions

<table>
<tr><th>reactive()</th><th>observe()</th><th>function() {...}</th></tr>
<tr><td>Callable</td><td>Not callable</td><td>Callable</td></tr>
<tr><td>Returns a value</td><td>No return value</td><td>Returns a value</td></tr>
<tr><td>Lazy</td><td>Eager</td><td>Lazy</td></tr>
<tr><td>Cached</td><td>N/A</td><td>Not cached</td></tr>
<tr><td>No side effects</td><td>Only for side effects</td><td>Side effects optional</td></tr>
</table>

---

## observeEvent and eventReactive

Every reactive expression or reactive value read by a `reactive()` or `observe()` block automatically becomes a *reactive dependency* of that reactive expression/observer.

`observeEvent` and `eventReactive` give us finer control.

```{r eval=FALSE}
observeEvent(input$save_button, {
  write.csv(movies_subset(), "movies.csv")
})
```

**"When the `save_button` button is clicked, write the value of `movie_subset` to disk."** (Don't write to disk automatically when `movie_subset` changes.)

---

## observeEvent and eventReactive

- observeEvent is for event handling
- eventReactive is for delayed computation

```{r eval=FALSE}
observeEvent(when_this_changes, {
  do_this
})

r <- eventReactive(when_this_changes, {
  recalculate_this
})
```

Use these functions when you want to **explicitly name your reactive dependencies**, as opposed to letting `reactive`/`observe` implicitly depend on anything they read.

---

## Exercise: CRAN logs

Open `apps/reactivity/cranlogs.R` and run it. This app has several problems:

- We get an error right off the bat--the plot is running before the user has specified any packages.
- Unless you're a very fast typist, typing package names will cause the cranlogs server to be queried with many incomplete queries.

Add an "Update" `actionButton` to the UI, and make sure nothing happens until it's clicked.

```{r echo=FALSE}
countdown::countdown_timer(60 * 5)
```

---

## Solution

See `apps/reactivity/cranlogs.solution.R`.

```{r eval=FALSE}
packages <- reactive({
  strsplit(input$packages, " *, *")[[1]]
})
```

becomes

```{r eval=FALSE}
packages <- eventReactive(input$update, {
  strsplit(input$packages, " *, *")[[1]]
})
```

---

## Review: reactiveValues

Read/write versions of `input`.

```{r eval=FALSE}
# Create
rv <- reactiveValues(x = 10)

# Read
rv$x

# Write
rv$x <- 20
```

Try not to use this to store _calculated_ values. But in some cases, it's unavoidable.

---

## Exercise: Counter

Open the file `apps/reactivity/counter.R`. It has three action buttons:

- Increment: Increase the value by 1
- Decrement: Decrease the value by 1
- Reset: Set the value to 0

Unfortunately, it doesn't work. See if you can implement the server side.

```{r echo=FALSE}
countdown::countdown_timer(60 * 10)
```

---

## Solution

See `apps/reactivity/counter.solution.R`.

```{r eval=FALSE}
rv <- reactiveValues(count = 0)

observeEvent(input$increment, {
  rv$count <- rv$count + 1
})
observeEvent(input$decrement, {
  rv$count <- rv$count - 1
})
observeEvent(input$reset, {
  rv$count <- 0
})

output$value <- renderText({
  rv$count
})
```

---

## When to use reactiveValues

_Don't_ use `reactiveValues` when you're calculating a value based on other values and calculations that are already available to you.

_Do_ use `reactiveValues` to store state that otherwise would be lost from your graph of reactive objects. For example:

- A calculation over the history of something reactive
```{r eval=FALSE}
observeEvent(input$add, {
  rv$total <- rv$total + input$x
})
```

(Or a more elegant way to do the same, using [hadley/shinySignals](https://github.com/hadley/shinySignals):)

```{r eval=FALSE}
total <- shinySignals::reducePast(reactive(input$x), `+`, 0)
```

---

## When to use reactiveValues

- Tracking which of several events happened most recently
```{r eval=FALSE}
observeEvent(input$editMode, {
  rv$mode <- "edit"
})

observeEvent(input$previewMode, {
  rv$mode <- "preview"
})

output$page <- renderUI({
  if (rv$mode == "edit") {
    ...
  } else if (rv$mode == "preview") {
    ...
  }
})
```

---

## When to use reactiveValues

- To change the usual rules of reactivity

Normally, as soon as reactive expressions are invalidated (before they have recalculated) they invalidate everyone downstream who depends on them.

But sometimes recalculating will end up giving us the same value as the previous anyway, and any downstream recalculations might have been wasted work. So:

```{r eval=FALSE}
dedupeReactive <- function(rexpr, priority = 10) {
  rv <- reactiveValues(value = NULL)
  
  observe({
    rv$value <- rexpr()  # TODO: Handle errors
  }, priority = priority)
  
  reactive(rv$value)
}
```

---

## Preventing reactivity with isolate()

Use `isolate` from inside a reactive expression or observer, to ignore the implicit reactivity of a piece of code. Wrap it around expressions or a whole code block.

```{r eval=FALSE}
# Updates every time input$x or input$y change
r1 <- reactive({
  input$x * input$y
})

# Updates only when input$x changes
r2 <- reactive({
  input$x * isolate(input$y)
})

# Never updates; it will always have its original value
r3 <- reactive({
  isolate({
    input$x * input$y
  })
})
```

---

## Preventing reactivity with isolate()

This used to be an extremely common pattern:

```{r eval=FALSE}
observe({
  # Take a reactive dependency on input$save_button
  input$save_button
  
  isolate({
    write.csv(movies_subset(), "movies.csv")
  })
})
```

Does this look familiar?

--

```{r eval=FALSE}
observeEvent(input$save_button, {
  write.csv(movies_subset(), "movies.csv")
})
```

---

## Checking preconditions with req()

#### Cancel the current output (or observer) if a condition isn't met.

**`req(input$text)`**  
Ensure the user has provided a value for the `"text"` input

**`req(input$button)`**  
Ensure the button has been pressed at least once

**`req(x %% 2 == 0)`**  
Ensure that x is an even number

**`req(FALSE)`**  
Unconditionally cancel the current reactive, observer, or output

---

## Checking preconditions with req()

`req(cond)` is similar to:
- `stopifnot(cond)`
- `if (!cond) stop()`
- `assertthat::assert_that(cond)`

but with these differences:
- Errors during output rendering show up with bold red text in the UI; `req` just makes the output blank
- Rather than verifying that `cond` is _true_, `req` verifies that `cond` is _truthy_ (see `?isTruthy`)
  - Feels unnatural to be so arbitrary and nebulous, but this definition is just too practical for UI programming

---

## Checking preconditions with req()

Most importantly, `req` is like an error in that it "infects" the downstream elements of the reactive graph (if a reactive throws an error, then any other reactive/observer/output that tries to access it will also throw an error).

We'll illustrate with an exercise:

---

## Exercise: Dynamic UI

Open `app/reactivity/dynamic.solution.R` and run it.

It has lots of errors in the browser and the R console--ignore those for the moment.

From the app, upload the `diamonds.csv` file found in the same directory. Now everything looks good.

See if you can figure out why these errors appear when the app first comes up, and how you can get them to go away (first without `req`, and then, if you have time and can figure out how, using `req`).

```{r echo=FALSE}
countdown::countdown_timer(7 * 60)
```

---

## Anti-solution

Open `apps/reactivity/dynamic.antisolution.R`.

This is how you used to have to do it: check for missing values yourself, and `return(NULL)`.

You had to do this in every reactive, observer, or output that could have a missing value, _plus_ all of the reactives, observers, and outputs that are downstream!

---

## Solution

Open `apps/reactivity/dynamic.solution.R`.

Now you can use `req` in the reactives, observers, and outputs that _directly_ use potentially-missing inputs, and everything downstream can just not worry about it.

---

## Time as a reactive source: invalidateLater

Besides using changing values as the impetus for reactive invalidation, we can also use the passage of time.

```{r eval=FALSE}
r <- reactive({
  Sys.time()
})

output$text <- renderPrint({
  r()
})
```

What will this produce?

---

## Time as a reactive source: invalidateLater

Besides using changing values as the impetus for reactive invalidation, we can also use the passage of time.

```{r eval=FALSE}
r <- reactive({
{{ invalidateLater(1000)}}
  Sys.time()
})

output$text <- renderPrint({
  r()
})
```

How about now?

---

## Near-realtime data sources

Use `reactiveFileReader` and `reactivePoll` to keep up-to-date with potentially changing data sources.

Winston and I will cover these functions in the Shiny Dashboards tutorial tomorrow.

---

## debounce and throttle

If a reactive value or expression changes too fast for downstream calculations to keep up, you can end up with a bad user experience (laggy experience, wasted work).

`debounce` and `throttle` take a reactive expression object as input, and return a _rate-limited_ version of that reactive expression.

```{r eval=FALSE}
# A reactive that updates as often as every 50 milliseconds
fast_reactive <- reactive({ ... })

# A reactive that updates no more often than every 2000 milliseconds
throttled_reactive <- throttle(fast_reactive, 2000)


# A reactive that doesn't update until fast_reactive has stopped
# changing for at least 1000 milliseconds
debounced_reactive <- throttle(fast_reactive, 1000)
```

Note that we pass `fast_reactive`, not `fast_reactive()`. Why?

---

## Exercise: points.R

Open and run `apps/reactivity/points.R`. Click on the plot a few times to create points. Notice the annoying laggy behavior--this is due to a (simulated) expensive summary output.

Use `debounce` or `throttle` to prevent the summary output from running so often.

```{r echo=FALSE}
countdown::countdown_timer(5 * 60)
```

---
class: middle

## End

